---
title: "Getting started with GQR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with GQR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(GQR)
```

## Purpose

GQR implements Generalised Q analysis as described by Dentinho, Kourtit and
Nijkamp (2023). It expands smaller sets of simple ranked or scored statements
into synthetic combined statements, constructs the W matrix, applies PCA, and
supports interpretation with statement and respondent regressions.

## Load the example data

The accessor returns a data frame directly:

```{r}
dat <- gqr_example_data("dummy_data")
str(dat)
```

The same object can be loaded through the conventional package-data interface:

```{r}
data("dummy_data", package = "GQR")
```

## Complete workflow

```{r}
fit <- gqr_analysis(
  data = dat,
  analysis_cols = paste0("Q", 1:9),
  id_col = "Respondent",
  dummy_mode = "all",
  n_components = 3,
  rotation = "varimax",
  respondent_regression = FALSE
)

fit
summary(fit)$variance
```

The full binary design has 512 rows because nine simple statements create
`2^9` subsets when the empty combination is retained.

## Grouped one-per-group design

The grouped design is appropriate when simple statements are alternatives
within separate questions or thematic blocks.

```{r}
groups <- data.frame(
  group = rep(c("Question 1", "Question 2", "Question 3"), each = 3),
  variable = paste0("Q", 1:9)
)

fit_grouped <- gqr_analysis(
  data = dat,
  analysis_cols = paste0("Q", 1:9),
  id_col = "Respondent",
  dummy_mode = "group_one_per",
  groups = groups,
  n_components = 3,
  rotation = "varimax",
  respondent_regression = FALSE
)

dim(fit_grouped$D)
fit_grouped$statement_regression$baselines
```

Exactly one statement is active in each group, producing `3 * 3 * 3 = 27`
synthetic combinations.

## Random design

```{r}
fit_random <- gqr_analysis(
  data = dat,
  analysis_cols = paste0("Q", 1:9),
  id_col = "Respondent",
  dummy_mode = "random",
  n_patterns = 200,
  prob = 0.5,
  seed = 101,
  n_components = 3,
  respondent_regression = FALSE
)

dim(fit_random$D)
```

## Step-by-step workflow

A dummy matrix is a binary design matrix: its columns are the original simple
statements and its rows are synthetic combinations. An entry of `1` includes a
statement in a combination; `0` excludes it.

```{r}
prepared <- gqr_prepare_data(
  dat,
  analysis_cols = paste0("Q", 1:9),
  id_col = "Respondent",
  transform = "standardise"
)

gqr_estimate_design(
  prepared$analysis_cols,
  mode = "all",
  n_respondents = nrow(prepared$data)
)

D <- gqr_generate_dummies(prepared$analysis_cols, mode = "all")
W <- gqr_make_w(prepared, D = D)
pca <- gqr_pca(W, n_components = 3, rotation = "varimax")
models <- gqr_regress_statements(pca, D)

head(models$coefficients)
```

## Gardening example

The bundled `gardening` object contains selected columns from the multilingual
Central and Eastern European gardening questionnaire dataset described by
Varga-Szilay et al. (2026). It is a package example, not a replacement for the
full published dataset.

```{r gardening, eval=FALSE}
data("gardening", package = "GQR")

analysis_columns <- c(
  "How_important_making_beautiful",
  "How_importnat_conservation",
  "How_important_making_money",
  "How_important_sel_supply",
  "Time_spend",
  "How_often_plant",
  "How_important_pesticide"
)

groups <- data.frame(
  group = c(rep("Motivation", 4), rep("Management", 3)),
  variable = analysis_columns
)

fit_gardening <- gqr_analysis(
  gardening,
  analysis_cols = analysis_columns,
  id_col = "ID",
  covariate_cols = c("Country_code", "Age_group", "Garden_size"),
  transform = "standardise",
  dummy_mode = "group_one_per",
  groups = groups,
  n_components = 4
)
```

## Graphical interface

```{r shiny, eval=FALSE}
run_gqr()
```

## References

Dentinho, T. P., Kourtit, K., & Nijkamp, P. (2023). Generalized Q analysis as a
new tool in social science research: A pedagogical introduction. *Eastern
Journal of European Studies*, **14**(2), 5–21.
doi:10.47743/ejes-2023-0201.

Varga-Szilay, Z., Šerić Jelaska, L., Vilumets, S., Barševskis, A., Benedek, K.,
Bevk, D., Jojczyk, A., Krištín, A., Růžičková, J., Veromann, E., Fetykó, K. G.,
Szövényi, G., & Pozsgai, G. (2026). A multilingual, multi-country dataset on
gardening and biodiversity awareness across Central and Eastern Europe.
*Scientific Data*. doi:10.1038/s41597-026-07887-9.
