---
title: "Methodology and interpretation in GQR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methodology and interpretation in GQR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(GQR)
```

## From traditional to Generalised Q analysis

Traditional Q methodology treats respondents as variables and statements as
observations so that shared subjective positions can be identified. Dentinho,
Kourtit and Nijkamp (2023) describe three practical difficulties: respondents
may be asked to rank too many statements, the statement count constrains the
number of non-redundant respondents, and naming components can be subjective.

Generalised Q analysis addresses these limitations by asking respondents to
rank or score smaller sets of simple statements, then systematically
recombining those responses into a larger synthetic statement space.

## Matrices

Let `V` have respondents in rows and simple statements in columns. Let `D` be a
binary matrix with synthetic combinations in rows and simple statements in
columns. GQR calculates:

```text
W = D %*% t(V)
```

Each value in `W` is a synthetic respondent evaluation of one combined
statement. Rows of `W` are combinations; columns are respondents.

## Design choices

### Full binary design

All subsets are included. This design is exhaustive but grows as `2^m`.

```{r}
gqr_estimate_design(18, mode = "all", n_respondents = 100)
```

### Grouped one-per-group design

This design represents combinations formed by selecting one alternative from
each question or thematic group. If group sizes are 4, 3 and 5, the design has
60 combinations rather than `2^12 = 4096`.

```{r}
variables <- paste0("S", 1:12)
groups <- list(
  Production = variables[1:4],
  Aesthetics = variables[5:7],
  Maintenance = variables[8:12]
)

gqr_estimate_design(
  variables,
  mode = "group_one_per",
  groups = groups
)
```

### Random design

Random Bernoulli combinations hold the number of rows fixed and can be used as
an exploratory approximation. Results should be checked across seeds or larger
samples of combinations.

## PCA orientation

GQR applies PCA to `W`, not to its transpose. Synthetic combinations are the
observations, so PCA scores describe combinations and PCA loadings describe
respondents. Varimax rotation is orthogonal and is intended to simplify the
component pattern without introducing correlations among components.

## Component interpretation

Statement regressions explain combination scores using the columns of `D`. In a
grouped one-per-group design, the dummies within each group sum to one, so one
statement per group is omitted as a baseline.

Respondent regressions explain respondent loadings with metadata. These models
can identify whether components are associated with demographic, geographic,
or other respondent characteristics.

Regression results are aids to interpretation, not automatic labels. Component
meaning should be assessed from the design, loadings, scores, coefficients,
substantive context, and robustness across analytical choices.

## Additivity and consistency

The W construction assumes that the evaluation of a combined statement equals
the sum of the constituent evaluations. This is the central modelling
assumption. Interactions, incompatibilities, or context effects among simple
statements are not represented automatically.

Dentinho et al. (2023) also describe consistency testing as an important future
and empirical use of the framework. Direct evaluations of selected combined
statements can, in principle, be compared with their synthetic values.

## Missing data and scaling

GQR can reject, mean-impute, or replace non-finite values with zero when
constructing W. The choice should be justified by the study design. Scaling can
strongly affect PCA when respondents or statements differ in variance.

## Reference

Dentinho, T. P., Kourtit, K., & Nijkamp, P. (2023). Generalized Q analysis as a
new tool in social science research: A pedagogical introduction. *Eastern
Journal of European Studies*, **14**(2), 5–21.
doi:10.47743/ejes-2023-0201.
