---
title: "Using the GQR Shiny application"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the GQR Shiny application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
library(GQR)
```

## Purpose of the application

The GQR Shiny application provides a graphical interface for the complete
Generalised Q analysis workflow. It is intended for users who prefer to prepare
their data, construct synthetic statement combinations, run principal component
analysis, and interpret the resulting components interactively.

Generalised Q analysis extends traditional Q methodology by constructing a
large set of synthetic combined statements from smaller groups of simple ranked
or scored statements. These combinations are represented by dummy variables:
binary indicators in which `1` means that a simple statement is included in a
given synthetic statement and `0` means that it is excluded. This approach can
reduce the burden placed on respondents, permit analyses with more respondents
than simple statements, and provide quantitative support for naming and
interpreting extracted components.

The application implements the method introduced by Dentinho, Kourtit and
Nijkamp (2023). The same workflow is also available through the ordinary R
functions documented in the other GQR vignettes.

<!-- SCREENSHOT PLACEHOLDER: Application landing page
Include a screenshot of the complete Home tab here.
Suggested filename: vignettes/figures/shiny-home.png
Suggested width: 100% of the vignette page.
-->

## Starting the application

Load GQR and launch the interface with:

```{r launch, eval=FALSE}
library(GQR)
run_gqr()
```

The application opens in the RStudio Viewer or in a web browser, depending on
your R and Shiny settings. It contains seven tabs:

1. **Home**: a short methodological introduction and information about the
   bundled gardening example;
2. **Data**: data loading, labelling, column roles, transformations, and groups;
3. **Dummies**: construction and inspection of the dummy matrix `D` and the
   synthetic evaluation matrix `W`;
4. **PCA**: principal component analysis, rotation, scree plot, and respondent
   loadings;
5. **Statement–Component Regression**: regression of combination scores on statement dummies;
6. **Component–Covariate Regression**: regression and graphical analysis of respondent loadings against
   selected covariates;
7. **Reproducible R script**: an executable record of the current analysis, including uploaded-input provenance and the GQR function calls needed to rerun it outside Shiny.

<!-- OPTIONAL GRAPHICAL ELEMENT: Workflow diagram
A simple diagram can be inserted here:
Home -> Data -> Dummies -> PCA -> Statement–Component Regression -> Component–Covariate Regression -> Reproducible R script
Suggested filename: vignettes/figures/shiny-workflow.png
-->

## Home tab

The Home tab introduces the methodological basis of GQR and explains the role
of dummy variables. It also identifies the published source of the bundled
gardening example. No analytical choices are made on this page.

The three workflow cards summarise the main stages:

- prepare the data;
- construct and analyse the synthetic combinations;
- interpret the components using plots and regressions.

After reviewing the introduction, continue to the **Data** tab.

## Data tab

The Data tab contains four collapsible sections. The choices made here are
frozen when **Next** is clicked. If you later change the selected variables,
transformations, or grouping, click **Next** again before continuing so that the
Dummies, PCA, and Component–Covariate Regression tabs use the revised settings.

<!-- SCREENSHOT PLACEHOLDER: Data tab overview
Include a screenshot showing the four collapsible Data sections.
Suggested filename: vignettes/figures/shiny-data-overview.png
-->

### 1. Data input

Data can be supplied in two ways.

#### Bundled example data

Select **Use built-in dataset**, then choose either:

- **Gardening**, which contains selected columns from the multilingual Central
  and Eastern European gardening questionnaire dataset published by
  Varga-Szilay et al. (2026); or
- **Dummy data**, a small synthetic example with nine analysis variables, one
  numeric covariate, and one two-level factor covariate.

#### External data

Use **Choose data file** to upload a file in one of the supported formats:

- CSV (`.csv`);
- RDS (`.rds`);
- R data (`.rda` or `.RData`).

For RDA or RData files, the selected object must be a data frame. Use **Open data
viewer** to inspect the imported object before proceeding.

If a CSV is read as a single column, check its delimiter and encoding before
continuing.

<!-- SCREENSHOT PLACEHOLDER: Data input
Include a screenshot showing the built-in dataset selector and file upload.
Suggested filename: vignettes/figures/shiny-data-input.png
-->

### 2. Column names and labels

Internal variable names come directly from the data frame and are used in the
analysis. Human-readable labels can be edited separately. These labels are
particularly useful when the original variable names are abbreviated or coded.

Edit the labels and click **Save labels**. Changing a label does not rename the
underlying column in the data frame.

<!-- SCREENSHOT PLACEHOLDER: Labels
Include a screenshot showing several internal names and edited labels.
Suggested filename: vignettes/figures/shiny-labels.png
-->

### 3. Column roles and transformations

Assign variables to two roles:

- **Columns for analysis** are the simple ranked or scored statements used to
  construct `D` and `W`. These columns must be numeric.
- **Covariate columns** contain respondent metadata used for filtering and for
  the respondent-level regressions in the Component–Covariate Regression tab.

For uploaded data, the application initially uses numeric columns as analysis
variables and non-numeric columns as covariates. For the bundled examples, the
default roles come from `gqr_example_roles()` so the graphical and scripted
interfaces start from the same definitions. In particular, **Dummy data** uses
`Q1`--`Q9` as analysis variables and automatically selects
`Numeric_covariate` and `Factor_covariate` as covariates. The Gardening example
uses its numeric questionnaire fields as analysis variables and its
non-numeric respondent descriptors (other than `ID`) as covariates. These are
only starting defaults and can be changed before clicking **Next**.

One transformation can be applied to all selected analysis columns:

- **No transformation**;
- **Standardise (z-score)**;
- **Normalise to [0,1]**;
- **Relative importance (sum = 1)**;
- **Shannon entropy**.

The appropriate transformation depends on the scale and interpretation of the
input values. Transformation changes the values used to construct `W`; it does
not modify the source file.

<!-- SCREENSHOT PLACEHOLDER: Roles and transformation
Include a screenshot showing selected analysis columns, covariates, and the
transformation choices.
Suggested filename: vignettes/figures/shiny-roles-transformations.png
-->

### 4. Groups of variables

Groups define sets of alternative statements. When groups are present, the
application constructs combinations containing exactly one statement from each
group. Without groups, the application uses the full binary design containing
all permitted inclusion/exclusion patterns.

Groups can be created in three ways.

#### Upload a grouping file

Upload a CSV with two columns named exactly `group` and `variable`:

```text
group,variable
Motivation,How_important_making_beautiful
Motivation,How_importnat_conservation
Motivation,How_important_making_money
Motivation,How_important_sel_supply
Management,Time_spend
Management,How_often_plant
Management,How_important_pesticide
```

Values in `variable` must exactly match the internal names of selected analysis
columns.

#### Create groups interactively

Enter a group name, click **Add group**, and assign analysis columns to it. A
statement should belong to only one group. Groups can be deleted or all grouping
can be cleared.

#### Use random grouping

The **Use random grouping** option randomly assigns the selected analysis
variables to the requested number of groups. This is a way to create a grouping
structure; it is not the same as the package-level random sampling of dummy
patterns available through `gqr_analysis(dummy_mode = "random")`.

<!-- SCREENSHOT PLACEHOLDER: Grouping interface
Include a screenshot with at least two groups and their assigned variables.
Suggested filename: vignettes/figures/shiny-groups.png
-->

When all settings are complete, click **Next**. This stores a snapshot of the
current data and opens the Dummies tab. If no groups are defined and the full
binary design is expected to be large, GQR first shows the number of synthetic
combinations and the approximate memory required by a fully materialised W
matrix. You can return to the grouping controls or explicitly continue without
groups.

## Dummies tab

The Dummies tab constructs the binary design matrix `D` and previews the
synthetic evaluation matrix `W`. Large dummy designs are generated in a
separate R process. A progress panel shows that the calculation is active, and
**Stop calculation** can terminate that process without terminating the main R
session.

Each column of `D` represents one analysis statement, and each row represents a
synthetic combination. For example, the pattern `(1, 0, 1)` includes the first
and third statements and excludes the second. These dummies are neither
artificial respondents nor missing-value codes.

The synthetic evaluation matrix is calculated as:

```text
W = D %*% t(V)
```

where `V` contains respondents in rows and selected analysis variables in
columns. Each value in `W` is therefore the sum of a respondent's scores for the
statements included in that synthetic combination.

<!-- SCREENSHOT PLACEHOLDER: Dummies tab introduction
Include a screenshot showing the dummy explanation and reported number of
combinations.
Suggested filename: vignettes/figures/shiny-dummies-introduction.png
-->

### Respondent filters

The collapsible filter panel can restrict respondents before PCA. Available
controls depend on the covariates selected on the Data tab:

- categorical covariates can be filtered by retained levels;
- numeric covariates can be filtered by ranges;
- respondent identifiers can be supplied as a comma-separated list.

These filters change the respondent columns used in the subsequent analysis but
do not change the dummy combinations.

### Dummy matrix preview

The dummy heatmap shows selected rows of `D`. Variables can be ordered:

- alphabetically;
- as they appeared on the Data tab;
- by group size.

Use this plot to verify that the intended combinations were generated. In a
grouped design, each row should contain exactly one active statement per group.

<!-- SCREENSHOT PLACEHOLDER: Dummy matrix D
Include a screenshot of the D heatmap. A grouped example is preferable because
the one-per-group structure is visually clear.
Suggested filename: vignettes/figures/shiny-d-matrix.png
-->

### W matrix preview

The W heatmap displays synthetic evaluations for combinations and respondents.
Use the slider to restrict the number of displayed combinations. For
responsiveness, the app calculates only the rows needed by this preview rather
than materialising the complete W matrix. Numeric values can be added to cells
for small previews.

<!-- SCREENSHOT PLACEHOLDER: W matrix
Include a screenshot of the W heatmap, preferably without numeric labels for a
large example and with labels for a small example.
Suggested filename: vignettes/figures/shiny-w-matrix.png
-->

Click **Proceed to calculations** after checking the design. Very large designs
produce an additional warning before continuing. The subsequent PCA runs in a
separate process and can also be stopped without terminating R.

## PCA tab

The PCA tab applies principal component analysis to the Generalised Q design.
Synthetic combinations are the observations, while respondents are the
variables represented by the PCA loadings.

For ordinary PCA, the app normally uses GQR's compact exact algorithm rather
than performing an SVD of the complete W matrix. Since `W = D V^T`, the rank of
W cannot exceed the number of original statements. The decomposition can
therefore be performed in the much smaller statement space and the retained
scores reconstructed afterwards. This is algebraically equivalent to ordinary
PCA on W, apart from floating-point error and arbitrary component signs.

The SPSS-style correlation option still uses a fully materialised W matrix. It
uses the same `psych::cor.smooth()` + `psych::principal()` workflow as the
original GQR application, including its Varimax component ordering and
regression-type scores. This preserves compatibility with earlier SPSS-oriented
results, but it has greater memory and computational requirements than the
compact ordinary-PCA engine.

<!-- SCREENSHOT PLACEHOLDER: PCA tab overview
Include a screenshot containing the scree plot and PCA controls.
Suggested filename: vignettes/figures/shiny-pca-overview.png
-->

### Explained variance and component selection

The scree plot shows the variance explained by successive components. Select
the number of retained components using the numeric control. The appropriate
number should be based on explained variance, the scree pattern, interpretability,
and the aims of the study rather than on a single automatic rule.

The scree plot can be downloaded as PNG or PDF.

### Rotation and PCA settings

Click **Run / update PCA** after changing the PCA settings. The progress panel
reports the current stage and provides **Stop calculation** at any time.

Available settings include:

- no rotation or Varimax rotation;
- centring;
- scaling;
- SPSS-style correlation-based scoring;
- mean imputation for missing values in SPSS mode.

Varimax is an orthogonal rotation intended to simplify component patterns. The
choice to centre, scale, or impute should be justified by the input data and the
study design.

### Loadings tables and scored W matrix

The app can open interactive tables containing unrotated and rotated respondent
loadings. Selected covariates are appended to these tables when available. The
scored W matrix can also be opened for inspection.

<!-- SCREENSHOT PLACEHOLDER: PCA loadings table
Include a screenshot of the rotated loadings table or the scored W viewer.
Suggested filename: vignettes/figures/shiny-pca-loadings.png
-->

## Statement–Component Regression tab

This tab supports statement-content interpretation. For each retained component,
its combination scores are regressed on the dummy variables that generated `W`.
The coefficients indicate which original statements are positively or
negatively associated with the component.

For a grouped one-per-group design, one variable from each group is omitted as a
baseline because all dummy variables within a group sum to one. Coefficients are
therefore interpreted relative to the omitted statement in the same group. The
heatmap retains these reference variables and displays their reference value as
zero. Variables that are constant or all zero because they were not activated
by the grouping are also retained and displayed as zero.

When groups are available, heatmap rows follow the group order rather than
alphabetical order. Horizontal lines separate groups, and group names are shown
to the right of the heatmap.

Choose between:

- standardised coefficients, which facilitate comparison among predictors; and
- unstandardised coefficients, which retain the original model scale.

The tab provides:

- a regression summary for the selected component;
- an interactive coefficient matrix;
- a coefficient heatmap across components;
- PNG and PDF downloads of the heatmap.

An essentially perfect fit may occur because combination scores are derived
from the same dummy design used as predictors. In that situation, coefficient
standard errors and p-values may be unreliable, while the coefficient pattern
can still support component interpretation.

<!-- SCREENSHOT PLACEHOLDER: PCA regression heatmap
Include a screenshot of the standardised coefficient heatmap.
Suggested filename: vignettes/figures/shiny-statement-regressions.png
-->

## Component–Covariate Regression tab

The Component–Covariate Regression tab relates respondent loadings to respondent-level covariates. It is
conceptually distinct from the Statement–Component Regression tab: the latter explains
combination scores using statements, whereas the Component–Covariate Regression tab explains respondent
loadings using metadata.

<!-- SCREENSHOT PLACEHOLDER: Component–Covariate Regression tab overview
Include a screenshot showing component and covariate selectors together with
one respondent-loading plot.
Suggested filename: vignettes/figures/shiny-output-overview.png
-->

### Filtering and model selection

Respondents can again be filtered by covariate values or identifiers. The
maximum number of displayed respondents can also be limited to keep plots
readable.

Select the component and covariates, then click **Run models**. Covariates with
no variation after filtering cannot be fitted. If no covariates were selected
on the Data tab, the app displays an informational message rather than a model.
Return to the Data tab, select covariates, and click **Next** again.

### Respondent loading map

The loading map plots respondents using two selected components. Points can be
coloured or otherwise grouped using available covariates. This view helps reveal
clusters, gradients, and unusual respondents.

<!-- SCREENSHOT PLACEHOLDER: Respondent loading map
Include a screenshot of a PC1-PC2 respondent map coloured by Country_code or
another categorical covariate.
Suggested filename: vignettes/figures/shiny-respondent-map.png
-->

### Regression tables and covariate plots

The regression table summarises associations between selected respondent
covariates and component loadings. A detailed model summary is available for an
individual component. The final plot displays component loadings against a
selected covariate, with controls appropriate to the covariate type.

These regressions should be treated as interpretive aids. Their validity still
depends on sample size, residual behaviour, coding choices, and the structure of
the respondent data.

<!-- SCREENSHOT PLACEHOLDER: Covariate relationship
Include one categorical boxplot or one continuous-covariate scatter plot from
the Component loadings by covariate section.
Suggested filename: vignettes/figures/shiny-covariate-plot.png
-->

## Worked example with the gardening data

The following sequence demonstrates the entire graphical workflow.

### Data settings

1. Open the Data tab.
2. Select **Use built-in dataset** and choose **Gardening**.
3. Retain the following analysis columns:

```text
How_important_making_beautiful
How_importnat_conservation
How_important_making_money
How_important_sel_supply
Time_spend
How_often_plant
How_important_pesticide
```

4. Select respondent covariates such as:

```text
Country_code
Age_group
Garden_size
Gardening_experience
```

5. Choose **Standardise (z-score)**.
6. Define two groups:
   - **Motivation**: the four variables beginning with `How_important_` or
     `How_importnat_`;
   - **Management**: `Time_spend`, `How_often_plant`, and
     `How_important_pesticide`.
7. Click **Next**.

This grouping produces `4 * 3 = 12` synthetic combinations because one statement
is selected from each group.

### Dummies and PCA

1. Inspect the dummy matrix and confirm that every row contains one Motivation
   variable and one Management variable.
2. Inspect the W heatmap.
3. Click **Proceed to calculations**.
4. Retain a small number of components initially, for example three or four.
5. Apply Varimax rotation and examine the scree plot and respondent loadings.

### Interpretation

1. In Statement–Component Regression, compare standardised coefficients across components.
2. Identify statements with the strongest positive and negative coefficients.
3. In Component–Covariate Regression, select respondent covariates and run the respondent regressions.
4. Compare the coefficient pattern with the respondent loading map and
   covariate plots.
5. Name components only after considering all of these outputs together with
   substantive knowledge of the survey.

<!-- SCREENSHOT PLACEHOLDER: Worked example montage
Optionally include a multi-panel figure containing the gardening D matrix,
scree plot, statement-coefficient heatmap, and respondent loading map.
Suggested filename: vignettes/figures/shiny-gardening-workflow.png
-->

## Common issues

### Changes made on the Data tab are not reflected elsewhere

Click **Next** again. The app deliberately freezes a data snapshot so that later
calculations do not change while controls are being edited.

### The design contains far more combinations than expected

Check whether groups were defined and whether every intended statement was
assigned. Without a valid grouping, the app uses the full binary design, whose
size grows as `2^m` for `m` statements.

### There are no covariate controls in the Component–Covariate Regression tab

Return to the Data tab, select one or more covariate columns, and click **Next**
again.

### A CSV appears as one column

The delimiter was probably not recognised. Save the file with a conventional
comma, semicolon, or tab delimiter, or use RDS/RDA instead.

### PCA reports insufficient variance

Some respondents may have constant values across all synthetic combinations,
or filtering may have retained too few informative respondents. Review the
selected statements, transformations, and filters.

### The app warns about a large number of combinations

Reduce the number of analysis statements, define meaningful groups, or perform
a random-design analysis through the non-Shiny API.

## Reproducing an analysis in R

GQR 0.1.2 adds a **Reproducible R script** tab that converts the current frozen Shiny analysis into executable R code. This is the preferred route from exploratory graphical analysis to a publication or archived script because the generated code calls the same exported `gqr_*()` functions used internally by the application.

The generated script records, where applicable:

- the bundled dataset name or original uploaded dataset filename;
- file size, dimensions and MD5 checksum for the dataset used in the analysis;
- all other datasets or grouping files uploaded during the session, including files later replaced and not used in the current analysis;
- selected analysis variables and covariates;
- transformation settings;
- the final group-to-variable mapping;
- dummy-matrix generation settings;
- respondent filters that were used when PCA was run;
- the PCA engine, component count, centring, scaling, rotation and SPSS-compatible settings;
- Statement–Component Regression after that tab has been used; and
- the most recent successfully fitted Component–Covariate Regression, including its selected components, covariates and respondent filters.

For uploaded files, Shiny itself stores a temporary path that ceases to be useful after the session ends. GQR therefore records the **original filename** and writes, for example:

```{r generated-upload-example, eval=FALSE}
data_file <- "survey_final.csv"
if (!file.exists(data_file)) stop("Required data file not found: ", data_file)

expected_data_md5 <- "..."
actual_data_md5 <- unname(tools::md5sum(data_file))
if (!identical(actual_data_md5, expected_data_md5)) {
  warning("The checksum of ", data_file,
          " differs from the file used in the Shiny analysis.")
}

dat <- GQR::gqr_read(data_file)
```

Place the original input file in the R working directory before rerunning the script, or edit `data_file` to its new location. Checksum verification can be disabled in the tab when it is not wanted.

Grouping files are treated slightly differently. Their filename and checksum are recorded for provenance, but the **final group mapping is written explicitly into the script**. Consequently, if the grouping was uploaded and then changed interactively, the generated script reproduces the edited groups rather than silently re-reading the older file.

Use **Copy script** to copy the current code to the clipboard or **Download .R** to save it directly. The **Append sessionInfo()** option adds the R/platform/package environment to the end of the script.

The script represents the analytical state that was actually used. For example, PCA parameters are captured when PCA is run rather than simply reading controls that may have been changed afterwards. Similarly, Component–Covariate Regression is included only after a model has been successfully fitted. Display-only settings such as heatmap appearance are not included because they do not change the analytical result.

## References

Dentinho, T. P., Kourtit, K., & Nijkamp, P. (2023). Generalized Q analysis as a
new tool in social science research: A pedagogical introduction. *Eastern
Journal of European Studies*, **14**(2), 5–21.
doi:10.47743/ejes-2023-0201.

Varga-Szilay, Z., Šerić Jelaska, L., Vilumets, S., Barševskis, A., Benedek, K.,
Bevk, D., Jojczyk, A., Krištín, A., Růžičková, J., Veromann, E., Fetykó, K. G.,
Szövényi, G., & Pozsgai, G. (2026). A multilingual, multi-country dataset on
gardening and biodiversity awareness across Central and Eastern Europe.
*Scientific Data*. doi:10.1038/s41597-026-07887-9.
